Practical Optional Types for Clojure

### 2015-07-05 formal-model.tex
Rip out most of section 2

We formalize our type system following~\citet{TF10}
(with differences highlighted in $\trdiff{\text{blue}}$).
The typing judgment 
$$
{\judgementrewrite   {\propenv}
              {\e{}} {\t{}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}
  {\ep{}}}
$$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}, and rewrites to \ep{}. We write 
{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
  or
{\judgement   {\propenv}
              {\e{}} {\t{}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
for expressions that rewrite to itself.

>>>>

We formalize our type system following~\citet{TF10}
(with differences highlighted in $\trdiff{\text{blue}}$).
The typing judgment 
$$
{\judgementrewrite   {\propenv}
              {\e{}} {\t{}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}
  {\ep{}}}
$$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}, and rewrites to \ep{}. We write 
{\judgementtworewrite{\propenv}{\e{}} {\t{}}{\ep{}} if we are only interested in the type
  or
{\judgementselfrewrite   {\propenv}
              {\e{}} {\t{}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
for expressions that rewrite to themselves.

### 2015-07-05 overview.tex
Rip out most of section 2

\begin{minted}{clj}
(greet nil)    ; Type Error:
               ;   Expected Str, given nil
               ; in: nil
\end{minted}

@@

\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, "
       (when i
         (apply str (repeat i (str "hello, "))))
       n "!"))
(greetings "Donald" 2)  ; :- Str
;=> "Hello, hello, hello, Donald!"
(greetings "Grace" nil) ; :- Str
;=> "Hello, Grace!"
\end{minted}

@@

\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]
  (str "Hello, "
       (apply str (repeat i (str "hello, ")))
       ; Type Error:
       ;   Expected Num, given (U nil Num).
       ; in: i
       n "!"))
\end{minted}

@@

\begin{minted}{clj}
  (fn [f] 
    (.getParent f)
    ; Type error:
    ;   Unresolved host interop: getParent
    )
\end{minted}

@@

\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)
  ; Type error:
  ;   Cannot call instance method 
  ;   java.io.File/getParent on type 
  ;   (U nil File).
  )
\end{minted}

@@

\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)
  ; Type error:
  ;   Expected Str, given (U nil Str).
  ; in: (.getParent f)
  )
\end{minted}

@@

\subsection{Multimethod}

A \emph{multimethod} is a kind of extensible function. It consists of
a \emph{dispatching function}, given with \clj{defmulti}, and a collection of \emph{methods} associated with
\emph{dispatch values}, each installed via \clj{defmethod}.

Recall the multimethod in figure~\ref{fig:ex1}.

Its dispatching function is
\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
respectively.

When called, the dispatching function is applied to the provided arguments to
yield a \emph{dispatching value}.
The \clj{isa?} function is then used to compare the current dispatching value to each method's dispatch value.
An unambiguous match leads to the corresponding method being applied to the arguments,
giving the final result.

The following Typed Clojure program is semantically identical to figure~\ref{fig:ex1}.

\begin{minted}{clj}
(ann pname [(U Str File) -> (U nil Str)])
(defn pname [x]
  (let [dfn  class ; dispatching function
        dval (dfn x)] ; dispatching value
    (cond
      (and (isa? dval String)       ; if
           (not (isa? dval File))) 
      (pname (new File x))          ; then

      (and (not (isa? dval String)) ; else if
           (isa? dval File))       
      (.getName x)                  ; then
      :else (throw (Exception. "No match")))))
\end{minted}

The \clj{isa?} predicate 
is a subclassing check for classes, otherwise
an equality check.

\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a :b) ;=> false
  (isa? String Object) ;=> true
  (isa? Object String) ;=> false
\end{minted}

@@

%A multimethod in Clojure is a function with a \emph{dispatch
%function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
%\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
%The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
%and \emph{dispatch function} \clj{class}, a function that
%returns the class of its argument or \clj{nil} if passed \clj{nil}.
%
%We can use {\clj{defmethod}} to install a method to \clj{path}.
%\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
%Now the dispatch table maps
%the \emph{dispatch value} \clj{String} to the function
%\clj{(fn [x] x)}. 
%We add another method
%which maps
%\clj{File} to the function
%\clj{(fn [x] (.getPath x))}
%in the dispatch table.
%\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
%
%After installing both methods, the call 
%$$
%\clj{(path (new File "dir/a"))}
%$$
%dispatches to the second method we installed because
%$$
%\clj{(isa? (class "dir/a") String)}
%$$
%is true, and finally returns 
%$$
%\clj{((fn [x] (.getPath x)) "dir/a")}.
%$$

We include the above sequence of definitions as \egref{example:rep}.

\begin{Code}
\begin{exmp}
\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
\label{example:rep}
\end{exmp}
\end{Code}

Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
type checks because \clj{:a} agrees with the parameter type \clj{Any},
but throws an error at runtime.

@@

\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
same length recursively call \clj{isa?} on the elements pairwise.
\begin{minted}{clojure}
  (isa? [Keyword Keyword] [Object Object]) ;=> true
\end{minted}

\egref{example:multidispatch}
simulates multiple dispatch by dispatching on
a vector containing the class of both arguments. \clj{open}
takes two arguments which can be strings or files and returns
a new file that concatenates their paths.

We call three different \clj{File} constructors, each known at compile-time
via type hints.
Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
except we update multiple bindings simultaneously.

Hash-maps with keyword keys play a major role in Clojure programming.
HMap types model the most common usages of keyword maps.

\begin{exmp}
\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
\label{example:decleaf}
\end{exmp}

%The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
%The function \clj{an-exp} is verified to return an \clj{Expr}.
%
Here \clj{defalias} defines \clj{Expr}, a type abbreviation
that describes the structure of a recursively-defined AST as a union of HMaps.
A quoted keyword in a type --- \clj{':if} --- is a singleton type that contains just the keyword.
A type that is a quoted map like \clj{'{:op ':if}} is a
HMap type with a fixed number of keyword entries of the specified types
known to be \emph{present},
zero entries known to absolutely be \emph{absent},
and an infinite number of \emph{unknown} entries entries.
Since only keyword keys are allowed, they do not require quoting.

\begin{exmp}
\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
\label{example:circleci}
\end{exmp}

\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
dissociating the raw \clj{:private-key} entry with \clj{dissoc}
and associating an encrypted private key
as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
The expression \clj{(:private-key kp)} shows that keywords are also 
functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
operation forgot to remove it.

\egref{example:absentkeys}
is like \egref{example:circleci}
except the \clj{:absent-keys} HMap option is used
instead of \clj{:complete?},
which takes a \emph{set literal} of keywords that do not appear in the map, written 
with \emph{\#}-prefixed braces.
The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
in the function body.

\begin{exmp}
\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
\label{example:absentkeys}
\end{exmp}

Since this example enforces that \clj{:private-key} must not appear
in a \clj{EncKeyPair}
Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
from the return value.
Now, however we could stash the raw private key in another entry
like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
without Typed Clojure noticing.

\paragraph{Branching on HMaps} Finally, testing on HMap properties
allows us to refine its type down branches. \clj{dec-map} takes an
\clj{Expr}, traverses to its nodes and decrements their values by \clj{dec}, then
builds the \clj{Expr} back up with the decremented nodes.

\begin{exmp}
\inputminted[linenos,firstnumber=1,firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
\label{example:decmap}
\end{exmp}

If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
we remove
the \clj{:do} and \clj{:const}
Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
and we are left with an \clj{:if} Expr.
On line 8,
we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
Line 9 know we can
remove the \clj{:const} Expr from the type of \clj{m} because it contradicts \clj{(= (:op m) :do)} being true,
and we know \clj{m} is a \clj{:do} Expr.
Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
so we are left with \clj{m} being a \clj{:const} Expr.

Section~\ref{sec:formalpaths} discusses how this automatic reasoning is achieved.

>>>>

\begin{minted}{clj}
(greet 
  nil ; Type Error:
      ;   Expected Str, given nil
  )
\end{minted}

@@

\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, "
       (when i  ; when i is a non-nil integer
         (apply str (repeat i "hello, ")))
       n "!"))
(greetings "Donald" 2)  ; :- Str
;=> "Hello, hello, hello, Donald!"
(greetings "Grace" nil) ; :- Str
;=> "Hello, Grace!"
\end{minted}

@@

\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]
  (str "Hello, "
       (apply str 
         (repeat 
           i ; Type Error:
             ;   Expected Int, given (U nil Int).
           "hello, "))
       n "!"))
\end{minted}

@@

\begin{minted}{clj}
  (fn [f] 
    (.getParent f) ; Type Error:
                   ;   Unresolved interop: getParent
    )
\end{minted}

@@

\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f) ; Type Error:
                 ;   Cannot call instance method 
                 ;   java.io.File/getParent on type 
                 ;   (U nil File).
  )
\end{minted}

@@

\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f) ; Type Error:
                 ;   Expected Str, given (U nil Str).
  )
\end{minted}

@@

@@

A \emph{multimethod} is a kind of extensible function. It combines
a \emph{dispatch function} with one or more \emph{methods}.
When invoked, the arguments are first supplied to the dispatch function, yielding
a \emph{dispatch value}. A method is then chosen
based on the dispatch value---this method is applied to the arguments
to finally return a value for the entire expression.

Recall the multimethod in figure~\ref{fig:ex1}, duplicated below.

%Its dispatching function is
%\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
%respectively.

The dispatch function \clj{class}---associated at multimethod creation with \clj{defmulti}---dictates 
whether the \clj{String} or \clj{File} method is chosen---both installed via \clj{defmethod}.

The multimethod dispatch rules involve
\clj{isa?}, a hybrid predicate which includes a subclassing check for classes and
an equality check for most other values.

\begin{minted}{clojure}
  (isa? String Object) ;=> true
  (isa? Object String) ;=> false
  ; keywords are interned symbols with a : prefix
  (isa? :a :a) ;=> true
  (isa? :a :b) ;=> false
\end{minted}

The current dispatch value and---in turn---each method's associated dispatch value
is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.

To demonstrate,
  \clj{(pname "STAINS/JELLY")}
chooses the \clj{String} method because
\clj{(isa? (class "STAINS/JELLY") String)}
is true and
\clj{(isa? (class "STAINS/JELLY") File)}
is false. Then the \clj{String} method body
\clj{(pname (new File "STAINS/JELLY"))}
chooses the \clj{File} method for opposite reasons,
resulting in 
\begin{minted}{clojure}
(.getName (new File "STAINS/JELLY"))  ; :- (U nil Str)
;=> "JELLY"
\end{minted}

%The following Typed Clojure program is semantically identical to figure~\ref{fig:ex1}.
%
%\begin{minted}{clj}
%(ann pname [(U Str File) -> (U nil Str)])
%(defn pname [x]
%  ; dispatch value calculated by applying dispatch
%  ; function `class` to argument `x`.
%  (cond
%    ; if (class x) subclasses String, but not File
%    (and (isa? (class x) String)
%         (not (isa? (class x) File)))
%    ; then choose the String method
%    (pname (new File x))
%
%    ; else if (class x) subclasses File, but not String
%    (and (isa? (class x) File)         
%         (not (isa? (class x) String)))
%    ; then choose the File method
%    (.getName x)
%    :else (throw (Exception. "No match"))))
%\end{minted}
%
%An unambiguous match leads to the corresponding method being applied to the arguments,
%giving the final result.

Dispatching on a non-class value is common, as well as specifying
a \clj{:default} method for when no methods are matched.

\begin{minted}{clojure}
(ann hi [Kw -> Str])
(defmulti hi identity)
(defmethod hi :en [_] "hello")
(defmethod hi :fr [_] "bonjour")
(defmethod hi :default [_] "um...")

(str (hi :en) " " (hi :fr)) ;=> "hello bonjour"
(hi :bocce)   ;=> "um..."
\end{minted}

The \clj{hi} multimethod dispatches on its arguments via the identity function.

%\subsection{Multimethods}
%
%A multimethod in Clojure is a function with a \emph{dispatch
%function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
%\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
%The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
%and \emph{dispatch function} \clj{class}, a function that
%returns the class of its argument or \clj{nil} if passed \clj{nil}.
%
%We can use {\clj{defmethod}} to install a method to \clj{path}.
%\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
%Now the dispatch table maps
%the \emph{dispatch value} \clj{String} to the function
%\clj{(fn [x] x)}. 
%We add another method
%which maps
%\clj{File} to the function
%\clj{(fn [x] (.getPath x))}
%in the dispatch table.
%\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
%
%After installing both methods, the call 
%$$
%\clj{(path (new File "dir/a"))}
%$$
%dispatches to the second method we installed because
%$$
%\clj{(isa? (class "dir/a") String)}
%$$
%is true, and finally returns 
%$$
%\clj{((fn [x] (.getPath x)) "dir/a")}.
%$$

%We include the above sequence of definitions as \egref{example:rep}.
%
%\begin{Code}
%\begin{exmp}
%\inputminted[firstline=5,lastline=10]{clojure}{code/demo/src/demo/rep.clj}
%\label{example:rep}
%\end{exmp}
%\end{Code}
%
%Typed Clojure does not predict if a runtime dispatch will be successful---\clj{(path :a)} 
%type checks because \clj{:a} agrees with the parameter type \clj{Any},
%but throws an error at runtime.

%\paragraph{Multiple dispatch} \clj{isa?} is special with vectors---vectors of the
%same length recursively call \clj{isa?} on the elements pairwise.
%\begin{minted}{clojure}
%  (isa? [Keyword Keyword] [Object Object]) ;=> true
%\end{minted}
%
%\inputminted[firstline=6,lastline=23]{clojure}{code/demo/src/demo/eg7.clj}
%
%\egref{example:multidispatch}
%simulates multiple dispatch by dispatching on
%a vector containing the class of both arguments. \clj{open}
%takes two arguments which can be strings or files and returns
%a new file that concatenates their paths.
%
%We call three different \clj{File} constructors, each known at compile-time
%via type hints.
%Multiple dispatch follows the same kind of reasoning as \egref{example:incmap},
%except we update multiple bindings simultaneously.

Immutable hash-maps with keyword keys play a major role in Clojure programming.

\begin{minted}{clojure}
(def breakfast {:en "waffles" :fr "croissants"})

breakfast ; :- (HMap :mandatory {:en Str :fr Str}
          ;          :complete? true)

(get breakfast :en)    ; :- Str
;=> "waffles"
; keywords are functions that look themselves up
(:fr breakfast)        ; :- Str
;=> "croissants"
(:bocce breakfast) ; :- nil
;=> nil
\end{minted}

HMap types model the most common usages of keyword maps.
The inferred type for \clj{breakfast} holds two kinds of information---the known entries \clj{:en}
and \clj{:fr} have string values (in \clj{:mandatory}) and that 
every other key is absent (in \clj{:complete?}).

HMap's default to being partially specified. The HMap shorthand annotating
\clj{lunch}---below---forgets 
information about absent keys, only providing \clj{:mandatory}
information.

\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})

(:en lunch)    ; :- Str
;=> "muffin"
(:fr lunch)    ; :- Str
;=> "baguette"
; Unknown lookups are now less accurate
(:bocce lunch) ; :- Any
;=> nil
\end{minted}

\begin{minted}{clojure}
(defalias RawKeyPair
  "Unencrypted keypair -- extra keys disallowed"
  (HMap :mandatory {:public-key RawKey,
                    :private-key RawKey},
        :complete? true))

(defalias EncKeyPair
  "Encrypted keypair -- extra keys disallowed"
  (HMap :mandatory {:public-key RawKey,
                    :enc-private-key EncKey},
        :complete? true))

(ann enc-keypair [RawKeyPair -> EncKeyPair])
(defn enc-keypair "Encrypt an unencrypted keypair"
  [{pk :private-key :as kp}] ; original map is kp
  (assoc 
    ; remove unencrypted private key
    (dissoc kp :private-key)
    ; add encrypted private key
    :enc-private-key (encrypt pk)))
\end{minted}
%\begin{exmp}
%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
%\label{example:circleci}
%\end{exmp}
Fully specified HMap's are used to
statically enforce the private key is not accidentally left in a supposedly
encrypted keypair,

\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad
  [{pk :private-key :as kp}]
  (assoc kp :enc-private-key (encrypt pk)))
\end{minted}

%\clj{enc-keypair} takes an unencrypted keypair and returns an encrypted keypair by
%dissociating the raw \clj{:private-key} entry with \clj{dissoc}
%and associating an encrypted private key
%as \clj{:enc-private-key} on an immutable map with \clj{assoc}.
%The expression \clj{(:private-key kp)} shows that keywords are also 
%functions that look themselves up in a map returning the associated value or \nil{} if the key is missing.
%Since \clj{EncKeyPair} is \clj{:complete?}, Typed Clojure enforces the return type
%does not contain an entry \clj{:private-key}, and would complain if the \clj{dissoc}
%operation forgot to remove it.

%\egref{example:absentkeys}
%is like \egref{example:circleci}
%except the \clj{:absent-keys} HMap option is used
%instead of \clj{:complete?},
%which takes a \emph{set literal} of keywords that do not appear in the map, written 
%with \emph{\#}-prefixed braces.
%The syntax \clj{(fn [{pkey :private-key, :as kp}] ...)}
%aliases \clj{kp} to the first argument and \clj{pkey} to \clj{(:private-key m)}
%in the function body.
%
%\begin{exmp}
%\inputminted[firstline=10,lastline=21]{clojure}{code/demo/src/demo/key2.clj}
%\label{example:absentkeys}
%\end{exmp}
%
%Since this example enforces that \clj{:private-key} must not appear
%in a \clj{EncKeyPair}
%Typed Clojure would still complain if we forgot to \clj{dissoc} \clj{:private-key}
%from the return value.
%Now, however we could stash the raw private key in another entry
%like \clj{:secret-key} which is not mentioned by the partial HMap \clj{EncKeyPair}
%without Typed Clojure noticing.

%\paragraph{Branching on HMaps} Finally, testing on HMap properties
%allows us to refine its type down branches. \clj{dec-map} takes an
%\clj{Expr}, traverses to its nodes and decrements their values by \clj{dec}, then
%builds the \clj{Expr} back up with the decremented nodes.
%
%\begin{exmp}
%\inputminted[linenos,firstnumber=1,firstline=15,lastline=27]{clojure}{code/demo/src/demo/hmap.clj}
%\label{example:decmap}
%\end{exmp}
%
%If we go down the then branch (line 4), since \clj{(= (:op m) :if)} is true
%we remove
%the \clj{:do} and \clj{:const}
%Expr's from the type of \clj{m} (because their respective \clj{:op} entries disagrees with \clj{(= (:op m) :if)})
%and we are left with an \clj{:if} Expr.
%On line 8,
%we instead strike out the \clj{:if} Expr since it contradicts \clj{(= (:op m) :if)} being false. 
%Line 9 know we can
%remove the \clj{:const} Expr from the type of \clj{m} because it contradicts \clj{(= (:op m) :do)} being true,
%and we know \clj{m} is a \clj{:do} Expr.
%Line 12 we strike out \clj{:do} because \clj{(= (:op m) :do)} is false,
%so we are left with \clj{m} being a \clj{:const} Expr.
%
%Section~\ref{sec:formalpaths} discusses how this automatic reasoning is achieved.

Partially specified HMap's with a common dispatch key
combine naturally with ad-hoc unions.

\begin{minted}{clojure}
(defalias Order   ; define type abbreviation
  "A meal order, tracking dessert quantities."
  (U '{:Meal ':lunch ; keyword singleton type
       :desserts Num}
     '{:Meal ':dinner :desserts Num}
     '{:Meal ':combo :meal1 Order :meal2 Order}))

(ann desserts [Order -> Int])
(defmulti desserts "Number of desserts per order."
          :Meal)  ; dispatch on :Meal entry
; map destructuring backwards, d is :desserts entry
(defmethod desserts :lunch [{d :desserts}] d)
(defmethod desserts :dinner [{d :desserts}] d)
(defmethod desserts :combo [{m1 :meal1 m2 :meal2]
  (+ (desserts m1) (desserts m2)))

(cutlery {:Meal :combo 
          :meal1 {:Meal :lunch :desserts 1}
          :meal2 {:Meal :dinner :desserts 2}})
;=> 3
\end{minted}

%\begin{exmp}
%\inputminted[firstline=6,lastline=13]{clojure}{code/demo/src/demo/hmap.clj}
%\label{example:decleaf}
%\end{exmp}
%
%The \clj{defn} macro defines a top-level function, with syntax like the typed \clj{fn}.
%The function \clj{an-exp} is verified to return an \clj{Expr}.
%
%Here \clj{defalias} defines \clj{Expr}, a type abbreviation
%that describes the structure of a recursively-defined AST as a union of HMaps.
%Keyword singleton types are quoted---\clj{':lunch}.
%A type that is a quoted map like \clj{'{:op ':if}} is a
%HMap type with a fixed number of keyword entries of the specified types
%known to be \emph{present},
%zero entries known to absolutely be \emph{absent},
%and an infinite number of \emph{unknown} entries entries.
%Since only keyword keys are allowed, they do not require quoting.

@@
